Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-20 demystify.tex
updates

    %       \item
    %       \item
    %       \item You do not need an MS (\autoref{sec:msrequirement}), and CS PhD takes about 5--7 years (\autoref{sec:time}). Compared to other countries, CS PhD in the US does not require an MS degree but has a longer PhD study time (\autoref{sec:non-us-differences}).
    %       \item
    %       \item
    %       \item
    %       %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
    %     \end{itemize}

@@

\chapterinfo{\emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries.}

@@

\section{Who is this document for?}

\chapter{How is Your Application Evaluated?}\label{sec:evalapps}

\chapterinfo{Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members.}

@@

Each application is assigned to about 3 adcom members, who will evaluate your profile and reach a consensus.  While the assigned reviewers are the main ones deciding your application, other faculty in the department will likely have access to your application and can provide inputs and opinions on your profile.

@@

@@

\section{Statement of Purpose (SOP)}\label{sec:sop}
\sectioninfo{\\emph{SOP is very important}. Write it in such a way that makes you \emph{stand out}.}

@@

\section{Your School and Grades}\label{sec:your-school}
\sectioninfo{\emph{High grades probably won't help much} (unless you're from a well-known school), but bad ones likely will raise concerns.}

@@

\section{CV/Resume}
The CV/Resume provides a summary of the applicant's achievements. Most schools require you to upload your CV with your application.
Prepare your CV in such a way that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, and Teaching Experience.

@@

\chapterinfo{Celebrating! Now it is your turn to evaluate the school!
Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now.}

@@

\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs (teaching assistantships) are provided by the department to help with classes (e.g., grading). RAs (research assistantships) are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but are very competitive.}

@@

For a full breakdown of how much a graduate student costs, see \autoref{sec:ra-cost}.
\sectioninfo{Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it.}

@@

\subsection{How much do \emph{I} cost?}\label{sec:ra-cost}

@@

\sectioninfo{One reason why you were rejected is because you applied to AI/ML, a super competitive field in recent years with many applicants.}

@@

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
No. In fact, students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.  However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{sec:your-school}).

@@

\section{How long to complete the CS PhD program?}\label{sec:time}

@@

\section{PhD in the US vs. Other Countries}\label{sec:non-us-differences}

>>>>

    %       \item
    %       \item
    %       \item (\autoref{sec:msrequirement}), and CS PhD takes abou (\autoref{sec:time}). (\autoref{sec:non-us-differences}).
    %       \item
    %       \item
    %       \item
    %       %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
    %     \end{itemize}

@@

\chapterinfo{\emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities in the US is not any harder than in other countries.}

@@

\section{How is Your Application Evaluated?}\label{sec:evalapps}

\sectioninfo{Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members.}

@@

Each application is assigned to about three adcom members, who will evaluate your profile and reach a consensus.  While the assigned reviewers are the main ones deciding your application, other faculty in the department will likely have access to your application and can provide inputs and opinions on your profile.

@@

\section{Who is this document for?}
\sectioninfo{This document is mainly intended for international students from smaller countries and domestic students from less well-known universities.}

@@

\section{Statement of Purpose (SOP)}\label{sec:sop}
\sectioninfo{\emph{SOP is very important}. Write it in such a way that makes you \emph{stand out}.}

@@

\section{Your School and Grades}\label{sec:your-school}
\sectioninfo{\emph{High grades probably won't help much} (unless you're from a very top and well-known school), but bad ones likely will raise concerns.}

@@

\section{CV/Resume}
\sectioninfo{Highlight and summarize major achievements, e.g., Publications, Programming Competition Awards, and Teaching Experience.}

The CV/Resume provides a summary of the applicant's achievements. Most schools require you to upload your CV with your application.
Prepare your CV in such a way that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, and Teaching Experience.

@@

\chapterinfo{Celebrate! Now it is your turn to evaluate the school!
Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now.}

@@

\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs are provided by the department to help with classes (e.g., grading). RAs are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but can be very competitive.}

@@

For a full breakdown of how much a graduate student costs, see \autoref{sec:ra-cost}.

@@

\subsection{How much do \emph{I} cost?}\label{sec:ra-cost}
\sectioninfo{Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it.}

@@

\sectioninfo{You aim too high, are overqualified, or even because you applied to AI/ML, a super competitive field in recent years with many applicants.}

@@

\section{Is an MS required for admission to PhD in CS?}\label{sec:msrequirement}
\sectioninfo{You do not need an MS to do PhD in CS.}

No. In fact, students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.  However, MS can help admission if it gives research experience or is from a more well-known school than your undergrad institution (\autoref{sec:your-school}).

@@

\section{How long to complete the CS PhD program?}\label{sec:time}
\sectioninfo{About 5--7 years in the US}

@@

\section{PhD in the US vs. Other Countries}\label{sec:non-us-differences}
\sectioninfo{Compared to other countries, CS PhD in the US does not require an MS degree but has a longer PhD study time.}
